---
title: "s3-1 Concordance of differentially expressed genes in the comparison of CB vs AM and that of Phd2KO vs wild-type AM"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   html_document:
     highlight: haddock
     toc: yes
     toc_depth: 2
     keep_md: yes
     fig_width: 5
     fig_height: 5
---

# Overview

Genes differentially expressed in the comparison of CB vs AM and that of Phd2KO vs wild-type AM are compared and their concordance will be analysed.

# Setup


```{r load_packages, message = FALSE, warning = FALSE}

project.dir <- "/fast/AG_Sugimoto/home/users/yoichiro/projects/20230620_Phd2KO_in_AM_and_CB"
renv::restore("/fast/AG_Sugimoto/home/users/yoichiro/projects/20230620_Phd2KO_in_AM_and_CB/R")

processors <- 7

temp <- sapply(list.files(
    file.path(project.dir, "R/functions"),
    full.names = TRUE
), source)

set.seed(1)

```

## The input and output files and the directories

```{r setup_directories}

annot.dir <- file.path(project.dir, "annotation/")
annot.ps.dir <- file.path(annot.dir, "mm39_annotation/processed_data/")

results.dir <- file.path(project.dir, "results")
s2.dir <- file.path(results.dir, "s2")
s3.dir <- file.path(results.dir, "s3")

create.dirs(c(
    s3.dir
))

```

# Import data

```{r import_master_table}

res.summary.dt <- fread(file.path(s2.dir, "all-de-results.csv"))

color.code <- c(
    "CB specific" = "#EE6677",
    "Others" = "gray80",
    "AM specific" = "#4477AA"
)

color.code.2 <- c(
    "PHD2KO_induced" = "#228833",
    "Others" = "gray80",
    "PHD2KO_repressed" = "#AA3377"
)

res.summary.dt[, `:=`(
    tissue_specific_genes = factor(tissue_specific_genes, levels = names(color.code)),
    tissue_specific_genes_Chang = factor(tissue_specific_genes_Chang, levels = names(color.code)),
    PHD2_regulated_in_AM = factor(PHD2_regulated_in_AM, levels = names(color.code.2)),
    PHD2_regulated_in_CB = factor(PHD2_regulated_in_CB, levels = names(color.code.2))
)]

```



# Definition of function

```{r def_func}

library("matrixStats")

plotMA <- function(res.summary.dt, data.postfix = "Tissue", data.vals = c("AM", "CB"), color.col, color.code, plot.title, log2fc_cap = 5){

    sl.dt <- copy(res.summary.dt)
    sl.dt[, `:=`(
        ave_TPM = rowMeans(cbind(
            get(paste0("TPM_", data.vals[1], "__", data.postfix)),
            get(paste0("TPM_", data.vals[2], "__", data.postfix))
        )),
        ave_log2fc = get(paste0("log2fc__", data.postfix))
    )]

    sl.dt[, `:=`(
        capped_log2fc = case_when(
            ave_log2fc > log2fc_cap ~ log2fc_cap,
            ave_log2fc < -log2fc_cap ~ -log2fc_cap,
            TRUE ~ ave_log2fc 
        ),
        capped_shape = case_when(
            abs(ave_log2fc) > log2fc_cap ~ 18,
            TRUE ~ 16
        )
    )]

    g1 <- ggplot(
        data = sl.dt[order(get(color.col) != "Others")],
        aes(
            x = ave_TPM,
            y = capped_log2fc,
            shape = as.character(capped_shape),
            color = get(color.col)
        )
    ) +

    geom_hline(yintercept = 0, color = "gray60") +
        geom_point() +
        scale_x_log10(
            breaks = scales::trans_breaks("log10", function(x) 10^x),
            labels = scales::trans_format("log10", scales::math_format(10^.x))
        ) +
        coord_cartesian(ylim = c(-log2fc_cap - 0.1, log2fc_cap + 0.1), xlim = c(1, 20000)) +
        scale_color_manual(values = color.code) +
        scale_shape_manual(values = c("16" = 16, "18" = 18), guide = "none") +
        theme(
            legend.position = "bottom",
            legend.title=element_blank(),
            aspect.ratio = 1
        ) +    
        xlab("Average mRNA abundance [TPM]") +
        ylab("log2fc in mRNA abudance") +
        ggtitle(plot.title)

    return(list(g = g1, sl.dt = sl.dt))

}


```

# Comparison of the effect of Phd2 KO in the CB and AM


```{r phd2ko_in_CB}

s1.sl.dt <- res.summary.dt[
(TPM_WT__CB_PHD2KO > 10 | TPM_PHD2KO__CB_PHD2KO > 10 |
 TPM_WT__AM_PHD2KO > 10 | TPM_PHD2KO__AM_PHD2KO > 10) &
!is.na(padj__CB_PHD2KO) & !is.na(padj__AM_PHD2KO)
]

export.base.cols <- c("gene_id", "gene_name", "gene_type")

s1.rawdata.dt <- copy(s1.sl.dt[, c(
    export.base.cols,
    "log2fc__CB_PHD2KO", "log2fc__AM_PHD2KO"
), with = FALSE])

setnames(
    s1.rawdata.dt,
    old = c("log2fc__CB_PHD2KO", "log2fc__AM_PHD2KO"),
    new = c("log2FC_by_Phd2ko_in_CB", "log2FC_by_Phd2ko_in_AM")
)

fwrite(s1.rawdata.dt, file.path(s3.dir, "Supplementary_Fig_1A.csv"))

g.phd <- s1.sl.dt %T>%
    {print(paste0("n = ", nrow(.)))} %>%
    ggplot(
        aes(
            x = log2fc__CB_PHD2KO,
            y = log2fc__AM_PHD2KO
        )
    ) +
    geom_hline(yintercept = 0, color = "gray60") +
    geom_vline(xintercept = 0, color = "gray60") +
    geom_point() +
    coord_cartesian(xlim = c(-10, 10), ylim = c(-10, 10)) +
    theme(
        aspect.ratio = 1
    ) +
    xlab("log2 FC by Phd2 KO in CB") +
    ylab("log2 FC by Phd2 KO in AM")



print(g.phd)

min.tpm.threshold <- 10

res.summary.dt[
(
    TPM_WT__CB_PHD2KO > min.tpm.threshold |
    TPM_PHD2KO__CB_PHD2KO > min.tpm.threshold |
    TPM_WT__AM_PHD2KO > min.tpm.threshold |
    TPM_PHD2KO__AM_PHD2KO > min.tpm.threshold
) &
!is.na(padj__CB_PHD2KO) & !is.na(padj__AM_PHD2KO)
]     %$%
    round(
        cor(
            log2fc__CB_PHD2KO, log2fc__AM_PHD2KO, method = "spearman"
        ),
        digits = 1
    )


```


# Analysis of the concordance of differentially expressed genes in the comparison of CB vs AM and that of Phd2KO vs wild-type AM


```{r concordance, fig.width = 7.5, fig.height = 6.5}

log2fc.cap <- 5

p.res.1 <- plotMA(
    res.summary.dt[
        !is.na(padj__Tissue) &
        !is.na(padj__AM_PHD2KO)
    ],
    data.postfix = "Tissue",
    data.vals = c("CB", "AM"),
    color.col = "PHD2_regulated_in_AM",
    color.code = color.code.2,
    plot.title = "CB vs AM\nPhd2 regulated genes in AM highlighted",
    log2fc_cap = log2fc.cap
)

fig1b.raw.dt <- copy(p.res.1$sl.dt)

fig1b.raw.dt <- fig1b.raw.dt[, c(
    export.base.cols,
    "TPM_AM__Tissue",
    "capped_log2fc",
    "PHD2_regulated_in_AM"
), with = FALSE]

setnames(
    fig1b.raw.dt,
    old = c("TPM_AM__Tissue", "capped_log2fc", "PHD2_regulated_in_AM"),
    new = c("TPM in AM (WT)", "log2FC (CB vs AM)", "class")
)

fwrite(fig1b.raw.dt, file.path(s3.dir, "Fig_1B.csv"))

p.res.1$sl.dt[order(padj__AM_PHD2KO, decreasing = TRUE)] %>%
    ggplot(
        aes(
            x = TPM_AM__Tissue,
            y = capped_log2fc,
            color = PHD2_regulated_in_AM
        )
    ) +
    geom_point() +
    geom_hline(yintercept = 0, color = "gray60") +
    ggrepel::geom_label_repel(aes(
                 label = ifelse(
                     tissue_specific_genes != "Others" &
                     PHD2_regulated_in_AM != "Others" &
                     sign(log2fc__Tissue) == sign(log2fc__AM_PHD2KO) &
                     gene_type == "protein_coding" &
                     (TPM_PHD2KO__AM_PHD2KO > 10 | TPM_WT__AM_PHD2KO > 10),
                     gene_name, NA                                                              
                 )
             ),
             max.overlaps = Inf, size = 5,
             xlim = c(NA, Inf), ylim = c(-NA, Inf),
             fontface = "italic"
             ) +
    scale_color_manual(values = color.code.2) +
    scale_x_log10(
        breaks = scales::trans_breaks("log10", function(x) 10^x)
    ) +
    coord_cartesian(xlim = c(1, 3000)) +
    theme(
        aspect.ratio = 0.8,
        legend.position = "bottom"
    )

fig1c.raw.dt <- copy(p.res.1$sl.dt)

fig1c.raw.dt <- fig1c.raw.dt[, c(
    export.base.cols,
    "ave_log2fc",
    "PHD2_regulated_in_AM"
), with = FALSE]

setnames(
    fig1c.raw.dt,
    old = c("ave_log2fc", "PHD2_regulated_in_AM"),
    new = c("log2FC (CB vs AM)", "class")
)

fwrite(fig1c.raw.dt, file.path(s3.dir, "Fig_1C.csv"))

ggplot(
    data = p.res.1$sl.dt,
    aes(
        x = PHD2_regulated_in_AM,
        y = ave_log2fc,
        fill = PHD2_regulated_in_AM
    )
) +
    geom_hline(yintercept = 0, color = "gray60") +
    geom_boxplot(alpha = 1, outlier.shape = NA) +
    coord_cartesian(ylim = c(-log2fc.cap, log2fc.cap)) +
    scale_fill_manual(values = color.code.2, guide = "none") +
    scale_x_discrete(guide = guide_axis(angle = 90)) +
    theme(
        aspect.ratio = 2.5
    ) +
    ylab("mRNA log2 FC (CB vs AM)")

p.res.1$sl.dt[!is.na(ave_log2fc), table(PHD2_regulated_in_AM)]

pval.tbl <- p.res.1$sl.dt %$%
    pairwise.wilcox.test(
        x = ave_log2fc,
        g = PHD2_regulated_in_AM,
        p.adjust.method = "none"
    )$p.value

p.adjust(c(
    pval.tbl["Others", "PHD2KO_induced"], pval.tbl["PHD2KO_repressed", "Others"]
))

```

# Confirmation of the results above using orthogonal data


```{r chang_data}

ggplot(
    data = res.summary.dt[
        !is.na(padj__Tissue) & !is.na(padj__Tissue_Chang)
    ],
    aes(
        x = log2fc__Tissue,
        y = log2fc__Tissue_Chang
    )
) +
    geom_point(alpha = 0.2, shape = 16) +
    theme(
        aspect.ratio = 1
    ) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(-5, 5)) +
    xlab("mRNA log2 FC (CB vs AM)") +
    ylab("mRNA log2 FC (CB vs AM) by Chang et al.")

res.summary.dt[
    !is.na(padj__Tissue) & !is.na(padj__Tissue_Chang)
] %$%
    cor.test(
        x = log2fc__Tissue, y = log2fc__Tissue_Chang,
        method = "pearson"
    )

res.summary.dt[
    !is.na(padj__Tissue) & !is.na(padj__Tissue_Chang)
] %$%
    cor.test(
        x = log2fc__Tissue, y = log2fc__Tissue_Chang,
        method = "spearman"
    )

p.res.2 <- plotMA(
    res.summary.dt[
        !is.na(padj__Tissue_Chang) & !is.na(padj__AM_PHD2KO)
    ],
    data.postfix = "Tissue_Chang",
    data.vals = c("CB", "AM"),
    color.col = "PHD2_regulated_in_AM",
    color.code = color.code.2,
    plot.title = "CB vs AM (Chang et al.)\nPhd2 regulated genes in AM highlighted",
    log2fc_cap = log2fc.cap
)

s1b.raw.dt <- copy(p.res.2$sl.dt)

s1b.raw.dt <- s1b.raw.dt[, c(
    export.base.cols,
    "ave_log2fc", "PHD2_regulated_in_AM"
), with = FALSE]

setnames(
s1b.raw.dt,
    old = c("ave_log2fc", "PHD2_regulated_in_AM"),
    new = c("log2FC (CB vs AM)", "class")
)

fwrite(s1b.raw.dt, file.path(s3.dir, "Supplementary_Fig_1B.csv"))

ggplot(
    data = p.res.2$sl.dt,
    aes(
        x = PHD2_regulated_in_AM,
        y = ave_log2fc,
        fill = PHD2_regulated_in_AM
    )
) +
    geom_hline(yintercept = 0, color = "gray60") +
    geom_boxplot(alpha = 1, outlier.shape = NA) +
    coord_cartesian(ylim = c(-log2fc.cap, log2fc.cap)) +
    scale_fill_manual(values = color.code.2, guide = "none") +
    scale_x_discrete(guide = guide_axis(angle = 90)) +
    theme(
        aspect.ratio = 2.5
    ) +
    ylab("mRNA log2 FC (CB vs AM)\n(Chang et al. data)")

p.res.2$sl.dt[!is.na(ave_log2fc), table(PHD2_regulated_in_AM)]

pval2.tbl <- p.res.2$sl.dt %$%
    pairwise.wilcox.test(
        x = ave_log2fc,
        g = PHD2_regulated_in_AM,
        p.adjust.method = "none"
    )$p.value

p.adjust(c(
    pval2.tbl["Others", "PHD2KO_induced"], pval2.tbl["PHD2KO_repressed", "Others"]
))


```


# Session information

```{r session info}

sessioninfo::session_info()

```
